
##           :The Analytics Edge
##           : Assignment Email Ham or Spam
##           : Reproducible notes following the assignment and lecture notes

##           : Anil Kumar
##           : IIT Madras



 [[source files available on GitHub](https://github.com/anilcs13m)]

 [[connect on linkedin]](https://in.linkedin.com/in/anilcs13m)]

## PRELIMINARIES

Load the library that are required in the assignment:
```{r load_packages, cache = FALSE, echo = TRUE, message = FALSE, warning = FALSE, tidy = FALSE}
library("tm")
library("SnowballC")

library("caTools")
library("rpart")
library("rpart.plot")
library("ROCR")
library("randomForest")

```


## INTRODUCTION

Nearly every email user has at some point encountered a "spam" email, which is an unsolicited message often advertising a
product, containing links to malware, or attempting to scam the recipient. Roughly 80-90% of more than 100 billion emails
sent each day are spam emails, most being sent from botnets of malware-infected computers. The remainder of emails are
called "ham" emails.

As a result of the huge number of spam emails being sent across the Internet each day, most email providers offer a spam
filter that automatically flags likely spam messages and separates them from the ham. Though these filters use a number
of techniques, most rely heavily on the analysis of the contents of an email via text analytics.

We will be looking into how to use the text of emails is spam or ham.

We will be extracting word frequencies from the text of the documents, and then integrating those
frequencies into predictive models.  

We are going to talk about __predictive coding__ -- an emerging use of text analytics in the area
of criminal justice.


### The _eDiscovery_ Problem 

we will build and evaluate a spam filter using a publicly available dataset first described in the 2006 conference paper 
"Spam Filtering with Naive Bayes -- Which Naive Bayes?" by V. Metsis, I. Androutsopoulos, and G. Paliouras. The "ham" messages 
in this dataset come from the inbox of former Enron Managing Director for Research Vincent Kaminski, one of the inboxes 
in the Enron Corpus. One source of spam messages in this dataset is the SpamAssassin corpus, which contains hand-labeled 
spam messages contributed by Internet users. The remaining spam was collected by Project Honey Pot, a project that collects spam messages and identifies spammers by publishing email address that humans would know not to contact but that bots might target with spam.
The full dataset we will use was constructed as roughly a 75/25 mix of the ham and spam messages.

### Predictive Coding

__Predictive coding__ is a new technique in which attorneys manually label some documents and then
use text analytics models trained on the manually labeled documents to predict which of the
remaining documents are responsive.


### The Data

The data set contains just two fields:

* __email__: the text of the email in question, 
* __responsive__: a binary (0/1) variable telling whether the email was spam.

The labels for these emails were made by attorneys as part of the 2010 text retrieval conference
legal track, a predictive coding competition.


## LOADING THE DATA

```{r loading_data}
emails = read.csv("emails.csv", stringsAsFactors=FALSE)
# How many of the emails are spam
table(emails$spam)
```

```{r check_df}
str(emails)
```

Let's look at a few examples (using the `strwrap()` function for easier-to-read formatting):
```{r example_email_1}
strwrap(emails$email[1])
```
We can see just by parsing through the first couple of lines that this is an email about a new working paper,
"_The Environmental Challenges and Opportunities in the Evolving North American Electricity Market_", 
released by the Commission for Environmental Cooperation, or CEC.    
While this certainly deals with electricity markets, it doesn't have to do with energy schedules
or bids, hence it is not _responsive_ to our query.
If we look at the value in the `responsive` variable for this email: 
```{r example_email_1_responsive}
emails$responsive[1]
```
we see that its value is __0__, as expected.

 
Let's check the second email: 
```{r example_email_2}
strwrap(emails$email[2])
```
The original message is actually very short, it just says _FYI_, and most of it is a forwarded message.
We have the list of recipients, and down at the very bottom is the message itself.
    "Attached is my report prepared on behalf of the California State auditor."
There is also an attached report.    
Our data set contains just the text of the emails and not the text of the attachments.
It turns out, as we might expect, that this attachment had to do with Enron's electricity bids in
California, and therefore this email is _responsive_ to our query.    
We can check this in the value of the `responsive` variable.
```{r example_email_2_responsive}
emails$responsive[2]
```
We see that that it is a __1__.

Let's look at the breakdown of the number of emails that are responsive to our query.
```{r table_of_responsive}
table(emails$responsive)
```
We see that the data set is unbalanced, with a relatively small proportion of emails responsive to
the query.  This is typical in predictive coding problems.


## CREATING A CORPUS

We will need to convert our tweets to a corpus for pre-processing. 
Various function in the `tm` package can be used to create a corpus in many different ways.    
We will create it from the `tweet` column of our data frame using two functions, `Corpus()` and `VectorSource()`.
We feed to this latter the `Tweets` _variable_ of the `tweets` _data frame_.

```{r create_corpus}
corpus <- Corpus(VectorSource(emails$email))
```

Let's take a look at corpus:
```{r check_corpus}
corpus
```


### Converting text to _lower case_

We use the `tm_map()` function which takes as

* its first argument the name of a __corpus__ and 
* as second argument a __function performing the transformation__ that we want to apply to the text.

To transform all text _to lower case_:

```{r process_tolower}
corpus <- tm_map(corpus, tolower)
```

```{r process_to_plain_text}
corpus <- tm_map(corpus, PlainTextDocument)
```

### Removing punctuation

```{r process_remove_punctuation}
corpus <- tm_map(corpus, removePunctuation)
```


### Removing _stop words_ 

Removing words can be done with the `removeWords` argument to the `tm_map()` function, with an
extra argument, _i.e._ what the stop words are that we want to remove, for which we simply 
use the list for english that is provided by the `tm` package.

We will remove all of these English stop words, but we will also remove the word "_apple_"
since all of these tweets have the word "_apple_" and it probably won't be very useful in our
prediction problem.

```{r process_remove_stopwords}
corpus <- tm_map(corpus, removeWords, stopwords("english"))
```


### Stemming

Lastly, we want to stem our document with the `stemDocument` argument.
```{r process_stemming}
corpus <- tm_map(corpus, stemDocument)
```

Now that we have gone through those four preprocessing steps, we can take a second look at the
first email in the corpus.
```{r check_corpus_after_cleaning}
strwrap(corpus[[1]])
```
It looks quite a bit different now.
It is a lot harder to read now that we removed all the stop words and punctuation and word stems,
but now the emails in this corpus are ready for our machine learning algorithms.


## BAG OF WORDS

### Create a _Document Term Matrix_

We are now ready to extract the __word frequencies__ to be used in our prediction problem.
The `tm` package provides a function called `DocumentTermMatrix()` that generates a __matrix__ where:

* the __rows__ correspond to __documents__, in our case tweets, and 
* the __columns__ correspond to __words__ in those tweets.

The values in the matrix are the number of times that word appears in each document.

```{r create_DTM}
DTM <- DocumentTermMatrix(corpus)
```

```{r check_DRM}
DTM
```
what we can see is that even though we have only __`r DTM$nrow`__ emails in the corpus, 
we have over __`r DTM$ncol`__ terms that showed up at least once, which is clearly too many
variables for the number of observations we have.  

So we want to remove the terms that don't appear too often in our data set.


### Remove sparse terms

Therefore let's remove some terms that don't appear very often. 
```{r remove_sparse_terms}
sparse_DTM <- removeSparseTerms(DTM, 0.97)
```

Now we can take a look at the summary statistics for the document-term matrix:
```{r check_lighter_DTM}
sparse_DTM
```
We can see that we have decreased the number of terms to __`r sparse_DTM$ncol`__, 
which is a much more reasonable number.


### Creating a Data Frame from the _DTM_

Let's convert the sparse matrix into a data frame that we will be able to use for our predictive models.
```{r convert_DTM_to_DF}
labeledTerms <- as.data.frame(as.matrix(sparse_DTM))
```

To make all variable names _R-friendly_ use:
```{r fix_variable_names}
colnames(labeledTerms) <- make.names(colnames(labeledTerms))
```

We also have to add back-in the outcome variable
```{r add_outcome_variable}
labeledTerms$responsive <- emails$responsive
```

```{r check_DF}
# str(labeledTerms)
```

The data frame contains an awful lot of variables, __`r ncol(labeledTerms)`__ in total, 
of which __`r ncol(labeledTerms)-1`__ are the frequencies of various words in the emails, 
and the last one is `responsive`, _i.e._ the outcome variable.


### Split data in training/testing sets

Lastly, let's split our data into a training set and a testing set, putting __70%__ of the data in
the __training__ set.

```{r split_train_test}
set.seed(144)

split <- sample.split(labeledTerms$responsive, SplitRatio = 0.7)

train <- subset(labeledTerms, split == TRUE)
test <- subset(labeledTerms, split == FALSE)
```


## BUILD A CART MODEL

Now we are ready to build the model, and we will build a simple _CART model_ using the __default parameters__.
A random forest would be another good choice from our toolset.  
```{r CART_model}
emailCART <- rpart(responsive ~ . , data = train, method = "class")
```

```{r CART_plot_tree, fig.width = 5, fig.height = 4}
prp(emailCART)
```
We see at the very top is the word _California_.   

* If `californ` appears at least twice in an email, we are going to take the right path and
  predict that a document is __responsive__.     
  It is somewhat unsurprising that California shows up, because we know that Enron had a heavy 
  involvement in the California energy markets.
* Further down the tree, we see a number of other terms that we could plausibly expect to be
  related to energy bids and energy scheduling, like `system`, `demand`, `bid`, and `gas`.
* Down at the bottom is `jeff`, which is perhaps a reference to Enron's CEO, Jeff Skillings, 
  who ended up actually being jailed for his involvement in the fraud at the company.


### Out-of-Sample Performance of the Model

Now that we have trained a model, we need to evaluate it on the test set.  
We build an object `predictCART` that has the predicted probabilities for each class from our CART model,
by using the `predict()` function on the model `emailCART` and the test data with `newdata = test`.
```{r model_CART_predict_test}
predictCART <- predict(emailCART, newdata = test)
```
This new object gives us the predicted probabilities on the test set.
We can look at the first 10 rows with 
```{r check_predictions}
predictCART[1:10, ]
```

* The __first__ column is the predicted probability of the document being __non-responsive__.
* The __second__ column is the predicted probability of the document being __responsive__.
* They sum to 1.

In our case we are interested in the predicted probability of the document being responsive
and it would be convenient to handle that as a separated variable.
```{r predict_prob_vector}
predictCART.prob <- predictCART[ , 2]
```
This new object contains our test set predicted probabilities.

We are interested in the __accuracy__ of our model on the test set, _i.e._ out-of-sample.
First we compute the _confusion matrix_:
```{r CART_confusion_matrix}
cmat_CART <- table(test$responsive, predictCART.prob >= 0.5)
cmat_CART 

accu_CART <- (cmat_CART[1,1] + cmat_CART[2,2])/sum(cmat_CART)
```
* __Overall accuracy__ of this _CART_ model is __`r round(accu_CART,4)`__    
* __Sensitivity__ = __TP rate__ = `r cmat_CART[2,2]` / `r sum(cmat_CART[2,])` = __`r round(cmat_CART[2,2]/sum(cmat_CART[2,]),3)`__
* __Specificity__ = __(1 - FP rate)__ = `r cmat_CART[1,1]` / `r sum(cmat_CART[1,])` = __`r round(cmat_CART[1,1]/sum(cmat_CART[1,]),3)`__
* __FP rate__ = `r cmat_CART[1,2]` / `r sum(cmat_CART[1,])` = __`r round(cmat_CART[1,2]/sum(cmat_CART[1,]),3)`__
* __FN rate__ = `r cmat_CART[2.1]` / `r sum(cmat_CART[2,])` = __`r round(cmat_CART[2,1]/sum(cmat_CART[2,]),3)`__


### Comparison with the _baseline model_

Let's compare this to a simple baseline model that __always predicts non-responsive__ (_i.e._ the
most common value of the dependent variable).   
To compute the accuracy of the baseline model, let's make a table of just the outcome variable `responsive`.
```{r baseline_accuracy}
cmat_baseline <- table(test$responsive)
cmat_baseline

accu_baseline <- max(cmat_baseline)/sum(cmat_baseline)
```
The accuracy of the baseline model is then __`r round(accu_baseline,4)`__.     
We see just a small improvement in accuracy using the CART model, which is a common case in
unbalanced data sets.

However, as in most document retrieval applications, there are __uneven costs for different types of errors__ here.   
Typically, a human will still have to manually review all of the predicted responsive documents to
make sure they are actually responsive.    
Therefore: 

* If we have a __false positive__, _i.e._ a non-responsive document labeled as responsive, 
  the mistake translates to a bit of additional work in the manual review process but no further 
  harm, since the manual review process will remove this erroneous result.    
* On the other hand, if we have a __false negative__, _i.e._ a responsive document labeled as non-responsive 
  by our model, we will miss the document entirely in our predictive coding process.   

Therefore, we are going to assign a higher cost to false negatives than to false positives, which
makes this a good time to look at other cut-offs on our ROC curve.


### ROC curve

Let's look at the ROC curve so we can understand the performance of our model at different cutoffs.   
To plot the ROC curve we use the `performance()` function to extract the _true positive rate_ 
and _false positive rate_.
```{r ROCR_prediction_object}
predROCR <- prediction(predictCART.prob, test$responsive)
perfROCR <- performance(predROCR, "tpr", "fpr")
```

We then plot the ROC curve, with the option that color-codes the different cutoff thresholds.
```{r get_ROC_2, fig.width = 6, fig.height = 5}
plot(perfROCR, colorize = TRUE, lwd = 4)
```
The best cutoff to select is entirely dependent on the costs assigned by the decision maker to
false positives and true positives.   
However, we do favor cutoffs that give us a __high sensitivity__, _i.e._ we want to identify a 
large number of the responsive documents.

Therefore a choice that might look promising could be in the part of the curve where it becomes
flatter (going towards the right), where we have a __true positive rate of around 70%__ (meaning
that we're getting about 70% of all the responsive documents), and 
a __false positive rate of about 20%__ (meaning that we are making mistakes and accidentally
identifying as responsive 20% of the non-responsive documents.)

Since, typically, the vast majority of documents are non-responsive, operating at this cutoff
would result in a large decrease in the amount of manual effort needed in the _eDiscovery_ process.   

From the blue color of the plot at this particular location we can infer that we are looking at a
threshold around maybe 0.15 or so, significantly lower than 0.5, which is definitely what we would
expect since we favor false positives to false negatives.


### _Area Under the Curve_ (AUC)

```{r AUC}
auc_CART <- as.numeric(performance(predROCR, "auc")@y.values)
```
The AUC of the _CART_ models is __`r round(auc_CART, 4)`__, which means that our model can
differentiate between a randomly selected responsive and non-responsive document 
about __`r round(100*auc_CART, 1)`%__ of the time.

---

